\scuba: iterative map-making with the Sub-Millimetre User Reduction Facility

### 2012-08-23 smurf_paper.tex
more work on the start - mostly the example bolometer data section

\title[SMURF: an iterative map-maker for \scuba]{The Sub-Millimetre User
Reduction Facility: an iterative map-maker for \scuba}

@@

There are three general styles of map-making that are relevant to
reducing bolomter data in the literature. The simplest ``direct
methods'' involve some basic processing of the data to remove as much
noise contamination as possible, (e.g., using baseline removal and
other simple filters), and then re-gridding these cleaned data into a
map. Such was the basic recipe for the baseline reduction of chopped
data from SCUBA-2's predecessor SCUBA
\citep{1998ASPC..145..216J,2000ASPC..216..559J}, and MAMBO, another
camera from the same generation. Generally speaking, such methods are
fast, although dependending on the science goals and noise properties
of the data, they may not achieve the best noise performance on the
angular scales of interest. A particular case of this method was
successfully used for the reduction of data from BOLOCAM
\citep[e.g.][]{laurent2005}, and its younger sibling AzTEC
\citep[e.g.][]{scott2008}, using ``principal component analysis''
(PCA) cleaning. A statistical black-box removes the most correlated
components of the bolometer signals, enabling the detection of
point-sources very close to the theoretical white-noise limits of the
detectors with reasonable computation times. However, PCA cleaning is
not a good generic solution for producing maps of extended structures,
since such sources produce correlated signals amongst many detectors,
and are removed by this procedure.

@@

A reasonable model for correlated noise in \scuba\ is a single
``common-mode'' signal seen by all of the bolometers. Iterative
estimation and removal of this signal significantly lowers the $1/f$
knee, without compromosing structures on angular scales larger than
the array footprint. Residual independent drifts at lower frequencies
are removed with an iterative high-pass filter. This strategy enables
SMURF to reduce data faster than they are taken, on single high-end
desktop computers, and it has successfully been used as part of a
real-time pipeline offering feedback to observers at the
telescope. The pipeline is also used to generate near-science grade
products using the JCMT Science Archive \citep{2011ASPC..442..203E}
hosted by the Canadian Astronomical Data Centre (CADC). Details of
the pipeline design are given in \citet{gibb2005} and
\citet{2008AN....329..295C}.

This paper is divided in the following way. We first describe the
properties of \scuba\ data and scan patterns in Section~\ref{sec:data}
to motivate the data model that we have adopted. Note that while most
of the analysis described here applies equally to all of the data
taken to-date with \scuba, the \scuba\ Shared-Risks Observing period
(S2SRO) which took place from 2010 February 11 to 2010 March 25 used
only a single subarray at each of the two observing wavelengths, and
had considerably different noise properties and poorer performance
than the fully-commissioned instrument in 2011 (a comparison is
provided in Section~\ref{sec:s2sro}. Next, the details of the SMURF
algorithm are given in Section~\ref{sec:algorithm}. The paper is
concluded in Section~\ref{sec:examples} with three detailed test cases
that span the majority of observation types likely to be undertaken
with \scuba, with an emphasis on the mitigation of divergence
problems, and characterizing the output maps: (i) Uranus, a bright,
compact source (Section~\ref{sec:point}); (ii) the Lockman Hole, a
blind survey of faint point-like source (Section~\ref{sec:cosmo}); and
(iii) the star-forming region M17, including bright, extended emission
(Section~\ref{sec:extended}). All of the data analyzed in this paper
are publicly available through the CADC SCUBA-2 raw-data queries page
for the dates and observation numbers given in the
text\footnote{\texttt{http://www3.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/
jcmt/search/scuba2}}. All of the analysis was performed using
\texttt{Starlink} version ... .

@@

The \scuba\ bolometer temperatures are read-out using superconducting
transition-edge sensors (TESs), with critical temperatures
$T_\mathrm{C}$ approximately 140 and 200\,mK at 850 and 450\,\micron,
respectively. The range about $T_\mathrm{C}$ in which the transition
from a superconductor to a normal material occurs is extremely small
(...typical value...?). The entire \scuba\ focal plane is held at a
\emph{base temperature} somewhat below $T_\mathrm{C}$. The TESs are
then biased up into the transition using two systems. First,
individual pixel heaters (using a current bias to apply the desired
Joule heating) are set to fixed values at the start of an observation
to compensate for most of the loading from the instrument and
atmosphere (in better weather, less compensation from the heaters is
required). Then, an electrical voltage bias is applied. As the
instrument observes, and there are tiny temperature fluctuations due
to astronomical sources and atmosphere, the impedence
changes. Thermoelectrical feedback helps to keep the TESs ``locked''
in the transition as this occurs; an increase in temperature increases
the impedence, and lowers the current and resulting heating
power. Changing magnetic fields resulting from these current changes
are then amplified using chains of superconducting quantum
interference devices (SQUIDs), and the larger output current is
finally digitized. Each detector has its own SQUID for the first-stage
of the amplification, but the remaining stages occur within a common
chain of SQUIDs for each of the 32 columns. All 40 rows are then
read-out in sequence at a row-visit frequency of about 12\,kHz. This
multiplexing scheme greatly reduces the number of readout and bias
lines to each detector, making such large-format bolometer array
designs feasible. It should also be noted that there is an additional
41st row of SQUID readouts that are not connected to TES. These ``dark
SQUIDs'' are used to track non-thermal noise sources that are common
to each column's amplifier chain.

@@

In general, bolometers have two predominant noise components: (i) the
effectively white thermal noise produced by the phonon and optical
loading on the absorbers (largely independent from bolometer to
bolometer); and (ii) and longer timescale drift resulting from slow
variations in the mean load (e.g. fridge, telescope, or atmospheric
changes), and also the readout electronics (often correlated amongst
many or all of the bolometers). As such, the spectrum corresponding to
a bolometer time-series exhibits a ``$1/f$'' shape below some knee
frequency, $f_\mathrm{knee}$, and is relatively flat at higher
frequencies. Other smaller effects may include line features in the
spectrum (a common culprit is pickup from AC mains, or microphonic
signals induced by mechanical vibrations), and finally a roll-off
towards the Nyquist frequency caused by the anti-aliasing filter that
precedes digitization.

@@

@@

In Fig.~\ref{fig:bolos_mix} we show sample time series from single
bolometers in each of the subarrays, as well as variations in the
mixing chamber temperature (as a proxy for the focal-plane
temperature), and the telescope pointing during a PONG scan.

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{bolos_point_mix.pdf}
\caption{A comparison between single bolometer time series in each
  band (top two plots), with the mixing chamber temperature and
  azimuth/elevation pointing offsets. There is a strong correlation
  between the bolometers and the roughly $\sim$25\,s oscillation in
  the fridge, but not with the telescope motion. Also note that the
  total power in this primary signal is similar at both 450 and
  850\,\micron, pointing to an internal rather than an external
  source. The grey signals overplotted in the top-two panels show the
  residual time-series after removing the common-mode signal. This
  operation shows: (i) that most of the low-frequency signal is common
  to all of the bolometers; and (ii) the non-correlated, and
  predominantly white noise at 450\,\micron\ is significantly larger
  than at 850\,\micron.}
\label{fig:bolos_mix}
\end{figure*}

Both bolometers share significant long-timescale structure
($\gsim10$\,s) that appears to be related to variations in the fridge
base temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, from $-0.04$\,pW to $+0.08$\,pW, are larger
than the $-0.03$\,pW to $+0.05$\,pW fluctuations at 850\,\micron. Such
behaviour might be expected if there is a comparable varying thermal
load from the fridge at each wavelength, but a larger contribution of
atmospheric variations through the 450\,\micron\ bandpass filters. We
also note that there is no obvious strong correlation between the
low-frequency signal structure, at either wavelength, with the
telescope motion.

The low-frequency signal component of the bolometer output is also
highly correlated amongst bolometers in the same subarray. We have
calculated a common-mode signal, $\mathbf{c}(t)$, as the average time
series of all the working bolometers. We then fit the amplitude of
$\mathbf{c}(t)$ at each wavelength to the signals shown in
Fig.~\ref{fig:bolos_mix} and remove it, yielding the grey residual
signals. These residuals are quite flat, although still with
noticeable variations. The white noise is also apparent, and larger at
450\,\micron\ as one would expect from the larger backgrounds compared
to 850\,\micron.

Next, we produce power spectral density (PSD) plots for several
representative bolometers in Fig.~\ref{fig:pspec}. To produce this
figure, we follow the convention that the PSD as a function of
frequency, $\mathbf{P}(f)$, is normalized such that the integral over
frequency gives the same variance as the time-series variance across
the full time series. In other words, given a bolometer signal
$\mathbf{b}(t)$,
%
\begin{equation}
\label{eq:psd}
\langle\mathbf{b}^2(t)\rangle = 2 \int_0^{f_\mathrm{N}} \mathbf{P}(f)
df .
\end{equation}
%
where we only integrate over the positive frequencies up to
$f_\mathrm{N}$, the Nyquist frequency, and the factor of 2 accounts
for the (equal) power that appears at negative frequencies in the
discrete Fourier Transform. The units of the PSD written in this form
are then pW$^2$\,Hz$^{-1}$; without the factor of 2 in front the
convention would be to write the units as pW$^2$\,s.

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{pspec.pdf}
\caption{Bolometer power spectral densities. Four representative
  bolometers have been selected from each sub-array, and the
  flat-fielded (but otherwise raw) PSDs are shown as coloured dotted
  lines (the blue signals are for the same time series as those shown
  in Fig.~\ref{fig:bolos_mix}). Horizontal dotted lines at $5 \times
  10^{-8}$ and $5 \times 10^{-9}$\,pW\,Hz$^{-1}$ at 450 and
  850\,\micron, respectively, give the approximate white-noise
  levels. There are broad line features in the PSDs at both
  wavelengths above $\sim$35\,Hz. From $\sim$50\,Hz up to the Nyquist
  frequency, the gradual roll-off of the anti-aliasing filter is also
  evident.  At lower frequencies, the bolometer signals exhibit clear
  $1/f$ knees at approximately 2 and 4\,Hz at 450 and
  850\,\micron. The solid black lines are the PSDs of the common-mode
  signals at each wavelength, and the solid coloured lines show the
  PSDs of the bolometers once the common-mode is removed. These
  residual signals have significantly lower $1/f$ knees, approximately
  0.4 and 2\,Hz at 450 and 850\,\micron. Finally, the dashed black
  lines show the spectral shape produced by a point source, assuming
  that the telescope is slewing at 200\,arcsec\,sec$^{-1}$ (typical
  for S2SRO); for reference, the top horizontal axis shows the
  conversion from frequency to angular scale. We can see that the line
  features may add significant noise on the scale of the PSF. We also
  see that, for this scan velocity, the low-frequency noise begins to
  dominate on scales of approximately 8.3 and 1.7\,arcmin; these
  scales are comparable to the array footprint. .7\,arcmin; these
  scales are comparable to the array footprint. \bf{Per:The array foot
  print can't both be 1.7 and 8.3 arc in unless you used 4 arrays at
  450 and only one at 850?}}
\label{fig:pspec}
\end{figure*}

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded data. At each wavelength, there is a clear
$1/f$ knee at a few Hz, followed by a predominantly white spectrum
punctuated by line features above $\gsim 30$\,Hz, and finally a
roll-off caused by the anti-aliasing filter above $\gsim 70$\,Hz. As
indicated in the previous section, the correlation between the
low-frequency components of the different bolometer signals is
large. The solid black lines in Fig.~\ref{fig:pspec} indicate the PSDs
of the common-modes $\mathbf{c}(t)$ at each wavelength, which
reproduce most of the low-frequency structure, as well as the
high-frequency line features. The $\mathbf{c}(t)$ otherwise drop
substantially below the individual bolometer PSDs at high-frequency,
as expected if the bolometers are dominated by un-correlated white
noise. Finally, we note that the amplitudes and slopes of the
$\mathbf{c}(t)$ are similar at both 450 and 850\,\micron. However,
since the 450\,\micron\ white noise is larger than at 850\,\micron\
(shown approximately by the horizontal dotted lines), the $1/f$ knee
occurs at a \emph{lower} frequency at 450\,\micron.

Next, the common-mode signals are fit to each bolometer time series
and removed as in Fig.~\ref{fig:bolos_mix}, and the resulting PSDs are
shown as solid coloured lines. In this example, the residual signals
have $1/f$ knees significantly lower than in the raw PSDs.

For reference, the top horizontal axis has been converted to angular
scale assuming a scan velocity of 200\,arcsec\,sec$^{-1}$ which was
typical of the S2SRO observations. The power spectra of the PSFs in
each band (arbitrary normalized to peak values of $10^{-6}$ and
$10^{-7}$\,pW$^2$\,Hz$^{-1}$ at 450 and 850\,\micron\ respectively)
are also showed as dashed lines for this assumed scan velocity showing
that the smallest features resolvable by the telescope may be slightly
affected by the excess noise in the line features. At the lower
frequency end, the $1/f$ noise dominates at scales $\gsim 2$\,arcmin
in the raw data and $\gsim 10$\,arcmin in the common-mode cleaned data
at 450\,\micron, and at scales $\gsim 1$\,arscmin in the raw data and
$\gsim 2$\,arcmin in the common-mode cleaned data at 850\,\micron.

@@

An additional noise source that appears to be significant in only a
small subset of the S2SRO data is magnetic field pickup. Since the
bolometer signals are ultimately detected through the amplification of
magnetic fields, any additional changing fields within the cryostat
will also be detected.

@@

%-------------------------------------------------
\subsection{Differences between the S2SRO and 2011 re-commissioned instrument}
\label{sec:s2sro}
%-------------------------------------------------

Since S2SRO the fridge performance has been greatly improved using NTD
thermometry to servo the temperature to a constant value, and with new
wiring adequate base temperatures can now be achieved. However, the
variability across the focal plane continues to be a problem now that
there is a full complement of subarrays (which still includes the two
subarrays used during S2SRO). Optimal tuning to achieve the greatest
overall array sensitivity is described in \citet{holland2012}.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first six modes from a principal component analysis,
  ordered by decreasing significance, of the combined 450 and
  850\,\micron\ time-series bolometer data. For each component, the
  top plot shows the time series of the eigenvector, normalized by its
  \rms. The bottom coloured panels show the eigenvalues for the
  bolometers at each wavelength (the amplitude of the eigenvector in
  the time series). For reference, both the mean, $\bar{\lambda}$, and
  \rms, $\lambda_\mathrm{rms}$, eigenvalues for the bolometers in each
  subarray are also shown. The first two components have comparable
  amplitudes, and dominate by at least a factor of 10 any other
  components of the time series. We suspect that they are caused by
  atmospheric variations (expected to be brighter at 450\,\micron),
  and fridge oscillations (expected to be comparable at each
  wavelength). However, the two eigenvectors do not necessarily map
  directly to these physical sources under this interpretation; rather
  they are two \emph{different} linear combinations of the underlying
  physical signals that give rise to statistically uncorrelated
  components. The third component accurately reproduces an apparent
  data-acquisition glitch that appears only at 450\,\micron\ (see
  Section ???). The fourth component is a ...\,Hz line feature only
  observed in column 15 of the 450\,\micron\ subarray. The fifth
  component shows a series of steps and glitches that appear
  predominantly along the bottom-left edge of the 850\,\micron\
  subarray where the bolometers are probably poorly biased. Finally,
  the sixth component is degenerating into a less significant, and
  more randomly scattered signal across the focal plane. Subsequent
  components follow this trend, and would generally be kept in a
  PCA-cleaning procedure.}
\label{fig:pca}
\end{figure*}

@@

The \model{SMO} model uses a rolling mean or median boxcar filter to
calculate the low-frequency component of the bolometer signals, which
are then removed. In other words, this is an alternative to the
high-pass filtering for which \model{FLT} is generally used. The
primary reason for developing this model was to make it more robust
against ringing near the ends of the time-series (since the FFT
assumes the signal is periodic, and there may be a large step between
the start and finish), or residual spikes (for which the median
filtering is particularly useful).

>>>>

\title[\scuba: map-making with SMURF]{\scuba: iterative map-making
with the Sub-Millimetre User Reduction Facility}

@@

There are three general styles of map-making that are relevant to
reducing bolomter data in the literature. The simplest ``direct
methods'' involve some basic processing of the data to remove as much
noise contamination as possible, (e.g., using baseline removal and
other simple filters), and then re-gridding these cleaned data into a
map. Such was the basic recipe for the reduction of chopped data from
SCUBA-2's predecessor SCUBA
\citep{1998ASPC..145..216J,2000ASPC..216..559J}, and MAMBO, another
camera from the same generation. Generally speaking, such methods are
fast, although dependending on the science goals and noise properties
of the data, they may not achieve the best noise performance on the
angular scales of interest. A particular case of this method was
successfully used for the reduction of data from BOLOCAM
\citep[e.g.][]{laurent2005}, and its younger sibling AzTEC
\citep[e.g.][]{scott2008}, using ``principal component analysis''
(PCA) cleaning. A statistical black-box removes the most correlated
components of the bolometer signals, enabling the detection of
point-sources very close to the theoretical white-noise limits of the
detectors with reasonable computation times. However, PCA cleaning is
not a good generic solution for producing maps of extended structures,
since such sources produce correlated signals amongst many detectors,
and are removed by this procedure.

@@

A reasonable model for correlated noise in \scuba\ is a single
``common-mode'' signal seen by all of the bolometers. Iterative
estimation and removal of this signal significantly lowers the $1/f$
knee, without compromosing structures on angular scales smaller than
the array footprint. Residual independent drifts at lower frequencies
are removed with an iterative high-pass filter. This strategy enables
SMURF to reduce data faster than they are taken, on single high-end
desktop computers, and it has successfully been used as part of a
real-time pipeline offering feedback to observers at the
telescope. The pipeline is also used to generate near-science grade
products using the JCMT Science Archive \citep{2011ASPC..442..203E}
hosted by the Canadian Astronomical Data Centre (CADC). Details of the
pipeline design are given in \citet{gibb2005} and
\citet{2008AN....329..295C}.

This paper is divided in the following way. We first describe the
properties of \scuba\ data and scan patterns in Section~\ref{sec:data}
to motivate the data model that we have adopted. Note that most of the
analysis described here applies equally to all of the data taken
to-date with \scuba, including during the Shared-Risks Observing
period (S2SRO) from 2010 February 11 to 2010 March 25 (during which
only single subarrays were available at each of the two observing
wavelengths). A comparison between the S2SRO and fully-commissioned
instrument is given in Section~\ref{sec:s2sro}. Next, the details of
the SMURF algorithm are given in Section~\ref{sec:algorithm}. The
paper is concluded in Section~\ref{sec:examples} with three detailed
test cases that span the majority of observation types likely to be
undertaken with \scuba, with an emphasis on the mitigation of
divergence problems, and characterizing the output maps: (i) Uranus, a
bright, compact source (Section~\ref{sec:point}); (ii) the Lockman
Hole, a blind survey of faint point-like source
(Section~\ref{sec:cosmo}); and (iii) the star-forming region M17,
including bright, extended emission (Section~\ref{sec:extended}). All
of the data analyzed in this paper are publicly available through the
CADC SCUBA-2 raw-data queries page for the dates and observation
numbers given in the
text\footnote{\texttt{http://www3.cadc-ccda.hia-iha.nrc-cnrc.gc.ca/
jcmt/search/scuba2}}. All of the analysis was performed using
\texttt{Starlink} version ... .

@@

The \scuba\ bolometer temperatures are read-out using superconducting
transition-edge sensors (TESs), with critical temperatures
$T_\mathrm{C}$ approximately 140 and 200\,mK at 850 and 450\,\micron,
respectively. The range about $T_\mathrm{C}$ in which the transition
from a superconductor to a normal material occurs is extremely small
(...typical value...?). The entire \scuba\ focal plane is held at a
\emph{base temperature} somewhat below $T_\mathrm{C}$. The TESs are
then biased up into the transition using two systems. First,
individual pixel heaters (using a current bias to apply the desired
Joule heating) are set to fixed values at the start of an observation
to compensate for most of the loading from the instrument and
atmosphere (in better weather, less compensation from the heaters is
required). Then, an electrical voltage bias is applied. As the
instrument observes, and there are tiny temperature fluctuations due
to astronomical sources and atmosphere, the impedence
changes. Thermoelectrical feedback helps to keep the TESs ``locked''
in the transition as this occurs; an increase in temperature increases
the impedence, and lowers the current and resulting heating
power. Changing magnetic fields resulting from these current changes
are then amplified using chains of superconducting quantum
interference devices (SQUIDs), and the larger output current is
finally digitized. Each detector has its own SQUID for the first-stage
of the amplification, but the remaining stages occur within a common
chain of SQUIDs for each of the 32 columns. All 40 rows are then
read-out in sequence at a row-visit frequency of about 12\,kHz. This
multiplexing scheme greatly reduces the number of readout and bias
lines to each detector, making such large-format bolometer array
designs feasible. Such a high sample rate is unnecessary to produce
maps, so the data are re-sampled to approximately 200\,Hz.  It should
also be noted that there is an additional 41st row of SQUID readouts
that are not connected to TES. These ``dark SQUIDs'' are used to track
non-thermal noise sources that are common to each column's amplifier
chain.

@@

In general, bolometers have two predominant noise components: (i) the
effectively white thermal noise produced by the phonon and optical
loading on the absorbers (largely independent from bolometer to
bolometer); and (ii) longer timescale drift resulting from slow
variations in the mean load (e.g. fridge, telescope, or atmospheric
changes), and also the readout electronics (often correlated amongst
many or all of the bolometers). As such, the spectrum corresponding to
a bolometer time-series exhibits a ``$1/f$'' shape below some knee
frequency, $f_\mathrm{knee}$, and is relatively flat at higher
frequencies. Other smaller effects may include line features in the
spectrum (common culprits are aliased high-frequency noise detected
during the multiplexed readout of the arrays as described in
Section~\ref{sec:bolos}, pickup from AC mains, or microphonic signals
induced by mechanical vibrations), and finally a roll-off towards the
Nyquist frequency caused by the anti-aliasing filter that precedes
digitisation.

@@

%-------------------------------------------------
\subsection{Differences between the S2SRO and 2011 fully-commissioned instrument}
\label{sec:s2sro}
%-------------------------------------------------

Since S2SRO the fridge performance has been greatly improved using NTD
thermometry to servo the temperature to a constant value, and with new
wiring, adequate base temperatures can now be achieved. However, the
gradient in detector properties across the focal plane continues to be
a problem now that there is a full complement of subarrays (which
still includes the two subarrays used during S2SRO). Optimal tuning to
achieve the greatest overall array sensitivity is described in
\citet{holland2012}.

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{bolos_point_mix_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{bolos_point_mix.pdf}
\caption{A comparison between single bolometer time series in each of
  the 450 and 850\,\micron\ bands with the mixing chamber temperature
  and azimuth/elevation pointing offsets, before and after upgrading
  the instrument. The grey signals overplotted in the top two panels
  show the residual time-series after removing the common-mode
  signals.  Left: data taken during the S2SRO period, observation 29
  on 20100313. There is a strong correlation between the bolometers
  and the roughly $\sim$25\,s oscillation in the fridge, but not with
  the telescope motion. Also note that the total power in this primary
  signal is similar at both 450 and 850\,\micron, pointing to an
  internal rather than an external source.  The nearly flat
  common-mode subtracted signals show: (i) that most of the
  low-frequency signal is common to all of the bolometers; and (ii)
  the non-correlated, and predominantly white noise at 450\,\micron\
  is significantly larger than at 850\,\micron. Right: data taken with
  the fully-commissioned instrument, observation 38 on
  20111112. Unlike the S2SRO data, there is no strong signal produced
  by variations in the fridge temperature. The 450\,\micron\ data have
  minimal correlated low-frequency noise (as evidenced by the lack of
  a difference between the raw and common-mode subtracted signals),
  although the 850\,\micron data show a common signal correlated with
  the telescope motion (note that this scan has a significantly larger
  amplitude than the S2SRO data set).}
\label{fig:bolos_mix}
\end{figure*}

\begin{figure*}
\centering
\includegraphics[width=0.49\linewidth]{pspec_s2sro.pdf}
\includegraphics[width=0.49\linewidth]{pspec.pdf}
\caption{Bolometer power spectral densities for the same two data sets
  (before and after upgrades) used in Fig.~\ref{fig:bolos_mix}. Four
  representative bolometers have been selected at each wavelength, and
  the flat-fielded (but otherwise raw) PSDs are shown as coloured
  dotted lines (the blue signals are for the same time series as those
  shown in Fig.~\ref{fig:bolos_mix}).  The solid black lines are the
  PSDs of the common-mode signals at each wavelength, and the solid
  coloured lines show the PSDs of the bolometers once the common-mode
  is removed.  Finally, the dashed black lines show the spectral shape
  produced by a point source given the scan speeds for the two
  observations (120\,arcsec\,s$^{-1}$ in the left-hand plot, and
  190\,arcsec\,s$^{-1}$ in the right-hand plot); for reference, the
  top horizontal axes shows the conversion from frequency to angular
  scale. Horizontal dotted lines at $2 \times 10^5$ and $5 \times
  10^4$\,pW\,Hz$^{-1}$ at 450 and 850\,\micron, respectively, are
  provided as a visual reference for the white-noise levels. In
  addition to $1/f$ and white noise components, and line features, all
  of the PSDs exhibit the gradual roll-off of the anti-aliasing filter
  just below the Nyquist frequency.  Left: for the S2SRO data there
  are broad line features in the PSDs at both wavelengths above
  $\sim$35\,Hz.  At lower frequencies, the bolometer signals exhibit
  clear $1/f$ knees at approximately 2 and 4\,Hz at 450 and
  850\,\micron. Common-mode subtraction removes most of the correlated
  fridge oscillation signal, lowering the $1/f$ knees to approximately
  0.4 and 2\,Hz (or angular scales of 1.7 and 8,3\,arcmin) at 450 and
  850\,\micron, respectively. Right: data from the fully-commissioned
  instrument tend to have lower $1/f$ knees, and fewer line
  features. The remaining low-frequency noise, now that the fridge
  oscillations have been removed, is considerably more independent
  from bolometer to bolometer than in the S2SRO data (note the larger
  spread in the dotted lines at low frequencies compared to the
  left-hand plot). Finally, this example shows how much the noise in
  the arrays can vary over time. Despite selecting bolometers from the
  same subarrays in each example, these particular 850\,\micron\ data
  from the fully-commissioned instrument are somewhat noisier.}
\label{fig:pspec}
\end{figure*}

@@

In Fig.~\ref{fig:bolos_mix} we show sample time series from single
bolometers in each of the 450 and 850\,micron\ focal planes, as well
as variations in the mixing chamber temperature (as a proxy for the
focal-plane temperature), and the telescope pointing, for two data
sets, before and after the upgrades that followed S2SRO.

In the S2SRO data (left panel), observation 29 on 20100313, both
bolometers share significant long-timescale structure ($\gsim10$\,s)
that appears to be related to variations in the fridge base
temperature, although the similarity is clearly greater at
850\,\micron. In this particular case, the total power in the
fluctuations at 450\,\micron, from $-0.04$\,pW to $+0.08$\,pW, are
larger than the $-0.03$\,pW to $+0.05$\,pW fluctuations at
850\,\micron. Such behaviour might be expected if there is a
comparable varying thermal load from the fridge at each wavelength,
but a larger contribution of atmospheric variations through the
450\,\micron\ bandpass filters. We also note that there is no obvious
strong correlation between the low-frequency signal structure, at
either wavelength, with the telescope motion.

The low-frequency signal component of the S2SRO bolometer output is
also highly correlated amongst bolometers in the same subarray. We
have calculated a common-mode signal, $\mathbf{c}(t)$, as the average
time series of all the working bolometers. We then fit the amplitude
of $\mathbf{c}(t)$ at each wavelength to the signals shown in
Fig.~\ref{fig:bolos_mix} and remove it, yielding the grey residual
signals. These residuals are quite flat, although still with
noticeable variations. The white noise is also apparent, and larger at
450\,\micron\ as one would expect from the larger backgrounds compared
to 850\,\micron.

Data from the fully commissioned instrument, observation 38 on
20111112, are shown in the right panel of
Fig.~\ref{fig:bolos_mix}. Having solved the fridge oscillation
problem, these data no longer exhibit a correlation with the mixing
chamber (note that variations are still seen in the mixing chamber; it
is likely that these are not observed at the focal plane itself, where
the NTD thermometry used in the servo loop is performed). There is
minimal correlated low-frequency noise in these 450\,\micron\ data
(note that there is almost no difference once the common-mode has been
subtarcted). The 850\,\micron\ channel, however, exhibits a
significant signal that is obviously correlated with the telescope
motion. This signal is correlated amongs many of the detectors, and
common-mode removal yields fairly flat residuals. Part of the reason
that this is seen here, and not in the S2SRO data set shown, is that
the amplitude of the scan pattern is somewhat larger. Most of this
``scan-synchronous'' noise is attributed to magnetic field pickup, as
described in Section~\ref{sec:magpickup}.

Next, we produce power spectral density (PSD) plots for several
representative bolometers from both focal planes, using the same two
data sets, in Fig.~\ref{fig:pspec}. To produce this figure, we follow
the convention that the PSD as a function of frequency,
$\mathbf{P}(f)$, is normalized such that the integral over frequency
gives the same variance as the time-series variance across the full
time series. In other words, given a bolometer signal $\mathbf{b}(t)$,
%
\begin{equation}
\label{eq:psd}
\langle\mathbf{b}^2(t)\rangle = 2 \int_0^{f_\mathrm{N}} \mathbf{P}(f)
df ,
\end{equation}
%
where we only integrate over the positive frequencies up to
$f_\mathrm{N}$, the Nyquist frequency, and the factor of 2 accounts
for the (equal) power that appears at negative frequencies in the
discrete Fourier Transform. The units of the PSD written in this form
are pW$^2$\,Hz$^{-1}$; without the factor of 2 in front the convention
would be to write the units as pW$^2$\,s to distinguish the two
normalisations.

The dotted coloured lines in Fig~\ref{fig:pspec} show the PSDs for
raw, though flat-fielded data. At each wavelength, and in both data
sets, there are clear $1/f$ knees at frequencies ranging from roughly
1 to 10\,Hz, followed by a predominantly white spectrum punctuated by
line features at higher frequencies, and finally a roll-off caused by
the anti-aliasing filter above $\gsim 70$\,Hz. As indicated in the
previous section, the correlation between the low-frequency components
of the different bolometer signals is large. The solid black lines in
Fig.~\ref{fig:pspec} indicate the PSDs of the common-modes
$\mathbf{c}(t)$ at each wavelength, which reproduce much of the
low-frequency structure, as well as some of the higher-frequency line
features. The $\mathbf{c}(t)$ otherwise drop substantially below the
individual bolometer PSDs at high-frequency, as expected if the
bolometers are dominated by un-correlated white noise. The common-mode
signals are fit to each bolometer time series and removed as in
Fig.~\ref{fig:bolos_mix}, and the resulting PSDs are shown as solid
coloured lines.  For reference, the top horizontal axes have been
converted to angular scale assuming the scan speeds of each
observation (120\,arcsec\,s$^{-1}$ in the left-hand plot, and
190\,arcsec\,s$^{-1}$ in the right-hand plot). The power spectra of
unresolved point-sources are also shown as dashed lines in each band
(arbitrarily normalized), showing that the smallest features
resolvable by the telescope are only minimally affected by the excess
noise in the line features.

In the S2SRO data (left-hand panel of Fig.~\ref{fig:pspec}), the
low-frequency noise is very correlated between the detectors (note the
tight scatter in the dotted colour lines, and their resemblance to the
common-mode), due to it being dominated by the fridge
oscillations. Common-mode subtraction is very effective, reducing the
$1/f$ knees in both wavelengths by about a factor of 5.

Raw data from the fully-commissioned instrument (right-hand panel of
Fig.~\ref{fig:pspec}) generally have noticeably lower $1/f$ knees
(with the exception of some of the noisier 850\,\micron\ bolometers in
this example) as compared with the S2SRO data. The residual
low-frequency noise is considerably less correlated amongst detectors
(larger spread in the dotted colour lines), leading to a less drastic
improvement upon common-mode removal. However, since the data are less
dominated by low-frequency noise to begin with, the signals generally
require less low-frequency filtering to produce maps, and therefore
retain larger-scale structures than with the S2SRO data. Furthermore,
there are generally fewer line features in the PSDs of bolometers in
the fully-commissioned instrument. Finally, these two data sets
illustrate how the white noise performance (NEFD) can vary. For this
particular data set, the 850\,\micron\ noise performance is
substantially worse than in the example from the S2SRO
period. Generally speaking such noise steps occur in only a subset of
the \scuba\ subarrays. However, given the 4-fold increase in available
working bolometers since S2SRO, and the improved low-frequency noise
performance, \scuba\ is now a significantly faster mapping instrument,
and preserves information up to larger angular scales.

@@

An additional noise source that appears to be significant in only a
small subset of the data (and moreso during the S2SRO period) is
magnetic field pickup. Since the bolometer signals are ultimately
detected through the amplification of magnetic fields, any additional
changing fields within the cryostat will also be detected.

@@

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{pca.pdf}
\caption{The first six modes from a principal component analysis,
  ordered by decreasing significance, of the combined 450 and
  850\,\micron\ time-series bolometer data. For each component, the
  top plot shows the time series of the eigenvector, normalized by its
  \rms. The bottom coloured panels show the eigenvalues for the
  bolometers at each wavelength (the amplitude of the eigenvector in
  the time series). For reference, both the mean, $\bar{\lambda}$, and
  \rms, $\lambda_\mathrm{rms}$, eigenvalues for the bolometers in each
  subarray are also shown. The first two components have comparable
  amplitudes, and dominate by at least a factor of 10 any other
  components of the time series. We suspect that they are caused by
  atmospheric variations (expected to be brighter at 450\,\micron),
  and fridge oscillations (expected to be comparable at each
  wavelength). However, the two eigenvectors do not necessarily map
  directly to these physical sources under this interpretation; rather
  they are two \emph{different} linear combinations of the underlying
  physical signals that give rise to statistically uncorrelated
  components. The third component accurately reproduces an apparent
  data-acquisition glitch that appears only at 450\,\micron. The
  fourth component is a periodic feature only observed in column 15 of
  the 450\,\micron\ subarray. The fifth component shows a series of
  steps and glitches that appear predominantly along the bottom-left
  edge of the 850\,\micron\ subarray where the bolometers are poorly
  biased. Finally, the sixth component is degenerating into a less
  significant, and more randomly scattered signal across the focal
  plane. Subsequent components follow this trend, and would generally
  be kept in a PCA-cleaning procedure.}
\label{fig:pca}
\end{figure*}

@@

The \model{SMO} model uses a rolling mean or median boxcar filter to
calculate the low-frequency component of the bolometer signals, which
are then removed. In other words, this is an alternative to the
high-pass filtering for which \model{FLT} is generally used. The
primary reason for developing this model was to make it more robust
against ringing near the ends of the time-series (since the FFT
assumes the signal is periodic, and there may be a large step between
the start and finish), or residual spikes (for which the median
filtering is particularly useful). However, the gap-filling and
continuity algorithms that we have employed (Section~\ref{sec:gaps})
successfully mitigate these problems, and the \model{FLT} model is
substantially faster,
